---
title: "Mock exam"
output: github_document
author: "Sebastian Veuskens"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preliminaries 
## Load libraries 

```{r}
library(lavaan)         # For Holzinger data 
library(GPArotation)    # For Factor rotations 
library(ltm)            # For Abortion data and Item response theory 
library(poLCA)          # For latent class analysis 
```

## Load data 

```{r}
cheating <- read.table("Old Exams/Mock_exam_2023_24/cheating.txt")
head(cheating)
summary(cheating) 
cor_mat <- cor(cheating)
heatmap(cor_mat, scaling='none')
```

```{r}
#################
#### Modify #####
#################
data <- cheating 
formula_ltm <- data~z1 
q <- 1                                                  # Specify the number of factors used in the model 
p <- ncol(data)
i <- 1                                                  # Item to investigate further
j <- 1                                                  # Factor to investigate further 
missing_patterns <- rbind(c(1, 0, 0, 1), c(1, 0, 1, 0))
#################

#### For dichotomous and polytomous (categorical) data (?), describes each item with frequencies #### 
dsc <- descript(data)
dsc$perc 
dsc$items 
## P-values that test independence between two items (H_0 == independent)
# This is Goodness of fit (see Latent trait models, marginal frequencies)
dsc$pw.ass              # Applies the Chi-squared test and second-order margins to check for positive correlation in the data (not in the model!)
plot(dsc)

ltm_model <- ltm(formula_ltm, IRT.param = TRUE) 
cbind(ltm_model$pattern$X, ltm_model$pattern$obs)
head(data)
```

1. There is 16 response patterns (all possible response patterns are observed)
2. FRAUD is the item with the highest proportion of no's 
3. (1, 1, 1, 1) was observed 2 times, (0, 0, 0, 0) was observed 207 times 
4. Item 2 and 4 and item 1 and 4 are not significantly correlated 

## 2

Both models assume a Bernoulli distribution for the likelihood, 
meaning the probability to receive a positive/negative answer for item i,
given the value of the underlying factor.

In addition, the prior distribution is the same for the factor 
(standard (multivariate) normal distribution). 

A main difference is the parametrization. 
The latent trait model follows the GLLVM framework and consists of 
a systematic component: $$ and a link function $$, in order to determine $\pi_i(y)$. 

A different parametrization is used for the 2-PL model: Instead of a linear combination + 
link function, the $\pi_i(y)$ is given by: 
Here, the discrimination parameters $\alpha_{i1}$ and $\beta_{i1}$ are equivalent and 
describe, how well an item separates between different values for the latent factor model.
However, $\alpha_{i0}$ and $\beta_{i0}$ differ: The second one indicates a difficulty parameter 
(the higher the more difficult), while the first one indicates easyness of an item. 

## 3

```{r}
#### Latent trait model ####
## Computation 
ltm_model <- ltm(formula_ltm, IRT.param = TRUE)         # IRT.param - NEGATIVE difficulty parameter -> indicates which parametrization to use: If TRUE, low value means low difficulty (easy)
ltm_model_rip <- ltm(formula_ltm, IRT.param=FALSE)
# From IRT.param = TRUE to IRT.param = False: 
coefs <- summary(ltm_model)$coefficients
-coefs[i, j] * coefs[i + p, j]                          # == summary(ltm_model_rip)$coefficients[i, j]
# Alternative: Rasch model -> All \beta_{i1} are equal to 1, constraint does this (p+1 indicates discrimination parameters)
# ltm_model <- rasch(data, IRT.param=TRUE, constraint=cbind(p+1, 1))
coef(ltm_model_rip)
```

1. The FRAUD is the most difficult one, according to the model. 
2. LIEEXAM, LIEPAPER, FRAUD, COPYEXAM 

## 5

```{r}
alpha_zero <- ltm_model_rip$coeff[,1]
alpha <- ltm_model_rip$coeff[,2]                        # Discrimination parameters 
# st_alpha <- alpha / sqrt(1 + sum(alpha^2))            # Somehow correlation between item x_i and latent variable y_j 
st_alpha <- alpha / sqrt(1 + alpha^2)                   # Somehow correlation between item x_i and latent variable y_j 
st_alpha 

coef(ltm_model, prob=TRUE, order=TRUE) 
```

1. The standardized alpha for LIEEXAM, LIEPAPER are relatively high, meaning 
there is a high correlation between the factor and a response. FRAUD is less, but still 
relatively strong correlated. COPYEXAM is less strong correlated, is dominated by random variation and
not the factor. 
2. Most Probable is COPYEXAM with 20% positive answer, while the other three items
have low positive answer rates between 2% and 5%. 

## 5

```{r}

#### Goodness of fit of model #### 
fit_model <- fitted(ltm_model)
fit_model

expected <- fit_model[,ncol(data) + 1]                  # Expected frequencies are in last column in fit_model, before are the items 
observed <- ltm_model$pattern$obs                       # The frequencies observed in the data 
patterns <- ltm_model$pattern$X                         # Each pattern that was observed 
n_pat <- nrow(patterns) 

cbind(patterns, observed, expected)

## Degrees of freedom 
df <- n_pat - p * (q+1) - 1
df
## Chi-square test for complete patterns 
chi_sq_ltm <- sum((expected - observed)^2 / expected)
pvalue_chisq <- 1 - pchisq(chi_sq_ltm, df)
pvalue_chisq

## Likelihood ratio test 
lr <- 2 * sum(observed * log(observed / expected))      # observed here is same as n in slides               
pvalue_lr <- 1 - pchisq(lr, df)
pvalue_lr
```

See above for numbers.

The tests are not very reliable since there is many patterns with 
very low (<5) expected observations. 

## 6

It is better to look at the margins, either second or third order margins 
and test their residuals. 

## 7

```{r}
factor_scores <- factor.scores(ltm_model, method='EAP') # EAP: expected apriori pattern - Exp means expected, not exponent or so 
factor_scores                                           # I think low scores for z1 etc. mean low ability/agreement for these individiual with that certain pattern 
plot(factor_scores$score.dat$z1)

resp_pattern <- factor_scores$score.dat[,1:p]
tot_scores <- apply(resp_pattern, 1, sum)                # Number of positive responses per pattern (sum of ones)  
tot_scores  
plot(tot_scores, factor_scores$score.dat$z1)
## Visualize 
plot(ltm_model, legend=TRUE, cx = "bottomright", xlab="Attitude toward/Difficulty", lwd=3, cex.main=1.5, cex.lab = 1.3, cex=1.1)
```

It measures mostly the attitude to cheat in exams, while the LIEEXAM and LIEPAPER 
items are much clearer separated by the factor than the other two items. 
Thus, it measures the keeness to ly about an exam or paper. 

## 5

```{r}

```